Structuring College Access

### 2025-02-13 index.qmd
working on prez

### Introduction {.hide}
#### Talent is everywhere, but structure allocates talent to unequal opportunities

@@

- Enrollment Planning Service (EPS): software recommends which Geomarkets to recruit and which high schools to visit
- Student Search Service: purchase the contact information of prospects that meet "search filter" criteria

**Research questions** (focus on metropolitan areas):

1. What is SES and racial variation between, within Geomarkets? How does this change over time?
1. What is SES and racial composition of included vs. excluded prospects, if student list purchases filter on particular Geomarkets?

@@

### Background{.hide}
#### Creating Market Segment Model and Geomarkets

Project origins

Market Segment Model

- Allocates students into four market segments based on score-sending behavior of 1980 SAT test-takers
- "local" student submits most SAT scores to institutions in their own Geomarket
- "An “in-state” student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions

Geomarkets

> Zemsky & Oedel (1983) [p. 4] write that,  [Often] the market boundaries match formal political and educational divisions, reflecting natural channels of communication. Each major metropolitan area is composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs.

# Theory
### Test map

@@

>>>>

### Introduction {.hide}

@@

- Enrollment Planning Service (est. 1984): software recommends which Geomarkets to recruit and which high schools to visit
- Student Search Service (est. 1972): purchase the contact information of prospects that meet "search filter" criteria

Research questions

1. What is SES, racial variation between, within Geomarkets? Over time?
1. What is SES and racial composition of included vs. excluded prospects, if student list purchases filter on particular Geomarkets?

Research interest: how quantification reproduces social stratification

@@

### Background{.hide}
#### Geomarkets and Market Segment Model created by @RN4982

**Market Segment Model**, based on student SAT score-sending behavior

- Each student-to-college score assigned to one of four categories:
  - Local (in Geomarket); in-state (diff Geomarket); regional; national
- Students sorted into 4 market segments: local, in-state, regional, national
  - "Local" students submit most SAT scores to local institutions
- *Market Segment Profile* (produced by Geomarket): 
  - How many students in each market segment
- *Institutional Profile*  (produced by Geomarket): 
  - How many students by market segement send scores to college *i*

<style>
/* Remove the triangle marker from summary elements */
details summary {
  list-style: none; /* In some browsers, this alone may suffice */
}

/* For WebKit-based browsers (Chrome, Safari, etc.) */
details summary::-webkit-details-marker {
  display: none;
}
</style>

<div style="margin: 0.5em 0 0 0;">
<details style="margin: 0; padding: 0;">
  <summary>**Geomarkets**</summary>
<div style="font-size: 0.80em; margin-bottom: 0;">
> Boundaries [often] match formal political and educational divisions, reflecting natural channels of communication. Each major metro area is composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs [@RN4982, p. 4].
</div>
</details>
</div>

- Meant to reproduce the recruiting territories of admissions officers
- @RN4982 create 143 Geomarkets in three regions
- Enrollment Planning Service (EPS) uses 304 Geomarkets

# Theory

### Scholarship on quantification {.hide}
#### Discussion of correlation by @RN4975[chapter 1]

**Correlation**, extent to which variables move together

- Predictive analytics in two steps: 
    1. Measure correlations in previous cases (training data)
    1. Apply training data correlations to predict future cases
- But correlations from training data are point-in-time snapshot
  - Observed correlations caused by enduring structural inequality
  - "Predicting the future on the basis of the past threatens to reify and reproduce existing inequalities" [@RN4794, p. 224]

<div style="margin: 0.5em 0 0 0;">
<details style="margin: 0; padding: 0;">
  <summary>**@RN4982[chapter 3]**</summary>
<div style="font-size: 0.85em; margin-bottom: 0;">
> Our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations” [@RN4982, p. 42].
</div>
</details>
</div>

- Which variables correlated with student market segment (training data)?
  - Parental education, family income, SAT score, educational aspirations
  - Conclusion: demand for higher education is a function of social origin
- Recommendation: selective colleges target high-SES Geomarkets

### Scholarship on quantification {.hide}
#### Discussion of homophily by @RN4975[chapter 2]

**Homophily**, birds of a feather flock together

- Network science assumes homophily result of voluntary action by actors
- But homophily programmed into algorithms that create connections
- Homophily in market research/geodemography
  - Based on snapshot of existing social stratification, match vertically categorized consumers to vertically categorized producers
  - Amplifies the effect of initial stratificaiton on subsequent stratification

<div style="margin: 0.5em 0 0 0;">
<details style="margin: 0; padding: 0;">
  <summary>**@RN4982[chapter 4]**</summary>
<div style="font-size: 0.85em; margin-bottom: 0;">
> The layering of collegiate competition is primarily a socioeconomic layering. The hierarchical structure of collegiate competition reflects the stratified social and economic dimensions of communities from which colleges draw their students. Competition among colleges, as admissions officers have told us for so long, is a matter of keeping company with one’s peers [@RN4982, p. 72].
</div>
</details>
</div>

- Conducted a social network analysis (training data):
  - 2-mode network: students (1) tie to colleges (2) by sending scores
  - Colleges *i* and *j* compete if many students send scores to both
  - Conclusion: Like colleges compete for like (social origin) students

<div style="margin: 0.5em 0 0 0;">
<details style="margin: 0; padding: 0;">
  <summary> Recommendation: target Geomarkets, schools targeted by peer colleges</summary>
<div style="font-size: 0.85em; margin-bottom: 0;">
> On occasion, senior spokespersons for the profession worry that students outside the main market areas remain forgotten and hence, unchallenged. Inevitably, the increasing competition for students, the expense of travel and mailings, and internal political constraints compel institutions to concentrate their efforts where they will do the most good. The result is a natural reinforcing of the basic socioeconomic patterns that gave shape in the first place to the structure of college choice [@RN4982, p. 44].
</div>
</details>
</div>

### Research questions and expectations{.hide}

**Research goals**

- Examine extent of social inequaltiy between Geomarkets (RQ1)
- Show how application of Geomarkets produces inequality (RQ2)

**Research questions**

1. SES and racial variation between, within Geomarkets? Change over time?

    - Geomarkets/Market Segment Model based on social stratification
    - US Racial segregation, correlation between class and race

1. SES and racial composition of included vs. excluded prospects, if "student list" purchases filter on particular Geomarkets?

    - Excluding Geomarkets that are poor and non-White associated with:
      - Disproportionate exclusion of first-generation students
      - Disproportionate exclusion of non-white students
      - Disproportionate exclusion of first-generation, non-white students

# Methods

### Methods{.hide}
#### Data

**RQ1: SES and racial variation over time between, within Geomarkets**

- Census tract-level data
  - 1980, 2000 Census; 2016-2020 American Community Survey (ACS)
  - Variables: race/Hispanic origin; median income, % poverty, % BA+
- Geomarket shapefiles obtained from 2012 blog post
  - Each zip code assigned to one Geomarket
  - Partial spatial assigns each Census tract to one or more Geomarkets

**RQ2: SES and racial composition of student list purchases**

- 414 student lists -- associated with 2.5 million prospects -- purchaseds by 14 public universities from 2016-2020
  - Each prospect has a home zip code and a HS zip code (point geoms)
- Analyze purchases that include all Geomarkets in a metro area
  - Show who would be excluded had search criteria exluded a Geomarket

### Methods{.hide}
#### Research design: quantitative case study

Cases are metro areas associated with multiple Geomarkets (n = 17)

<div style="margin-top: 0.8em;"></div>

Case selection

- Initial case selection based on availability of student list purchases (RQ2)
  - Prefer purchases that filter on test score, HS class, GPA, but not on student preferences (e.g., college size)
    - Ideal: purchases differ on test score range, but same on other filters
  - For some metros, don't have purchases that include all Geomarkets
- Final set of cases for manuscript main text (3 metro areas)
  - TBD; unsure about criteria
  - Conducted analyses for larger sample of metros

Analyses (RQ1 and RQ2)

- Descriptive statistics for between Geomarket variation
- Interactive maps for within Geomarket variation

### Test map{.hide}

@@

**RQ1: SES and racial variation over time between, within Geomarkets**

- Substantial inequality between and within Geomarkets
- In 1980, Black people are usually concentrated in poorest Geomarket
  - In some metro areas, this persists (e.g., Cleveland)
  - In some metro areas,  (e.g., Atlanta)

**RQ2: SES and racial composition of student list purchases**

- Poor, non-white Geomarkets produce disproportionate share of Black, Hispanic students
- Poor, non-white Geomarkets produce disproportionate share of first-generation studnets
- Purchased student profiles by race X parental education
  - Asian students from low-income Geomarkets tend to be first-gen compared to Asain students from rich Geomarkets
  - Same result for Hispanic students
